The unphysicality of Hilbert spaces

### 2023-06-23 Articles/2023-HilbertQuantum/HilbertQuantum.tex
Adding appendix on Schwartz spaces

Recall that an inner product space $V$ is complete if every Cauchy sequence in $V$ converges in $V$. This definition is difficult to characterize physically, therefore we use an equivalent condition: an inner product space $V$ is complete is complete if and only if all absolutely convergent series converge in $V$ (see theorem 13.8 in \cite{roman_2008}). This reframes the requirement in terms of convergence of linear combinations, and therefore superpositions.

We now show that if we have an unbounded operator, like position, completeness requires the existence of pure states for which

If we are only interested in observable with finitely many possible outcomes, we can use finite dimensional spaces. But as soon we have discrete quantities with arbitrarily large values, such as the number of particles, or quantities that are continuous, like position and momentum, we need infinite dimensional spaces. If we allow infinite dimensional spaces, we need some way of controlling what infinite linear combinations are allowed or not allowed. Cauchy completeness is one such way, but if that is not physical, what should we use?

We saw that the main problem is having expectation values that diverge, therefore the naive idea would be to require that the expectation value for all possible variables converge. This would require the expectation for $e^x$, $e^{e^x}$, $e^{e^{e^x}}$ and so on to be finite as well, which seems rather extreme. But what is a reasonable point to stop? If the idea is that we need to be able to reconstruct the state from expectation values, we should note that bounded distributions can be recovered from all the moments of the distribution, from the expectation of $x^n$ for every finite $n$. It seems reasonable, then, to require the expectation of all polynomial of position and momentum to be finite.

In quantum mechanics, this means requiring all expectations of $X^nP^m = -\imath \hbar x^n\partial_x ^m$ to be finite. This restricts our space to those functions that are infinitely smooth (i.e. all derivative exist) and decrease very rapidly as $x$ goes to infinity: the Schwartz functions. Requiring finite expectation for all polynomials of position and momentum, then, does not give us the usual $L^2$ Hilbert space, it gives us the Schwartz space. Is this a better candidate?

First of all, the Schwartz space is a subspace of the $L^2$ Hilbert space taken as an inner product space. That is, all Schwartz functions are elements of the Hilbert space, they are closed under linear combinations and have the same inner product. It is not, however, complete, which is what we want. Moreover, it is a dense subspace, meaning that any element of the Hilbert space can always be approximated by a Schwartz function with an arbitrary level of precision. Or, equivalently, that once a map is defined over the Schwartz space, the map is uniquely extended over the Hilbert space. This is a good hint that we have thrown away only unphysical elements.

If we restrict ourselves to the Schwartz space, then, the only unitary transformations we are interested in are those that map Schwartz function to Schwartz functions. The change of coordinate we defined in the previous section, then, is ruled out. Limits in terms of Cauchy sequences are not guaranteed, but limits in terms of expectation values are guaranteed: if we have a sequence of states such that the limit of the expectation converge, then we are guaranteed that the limit is also a Schwartz function.

The Schwartz space has also a very important property: it is the only space that is closed under the Fourier transform. That is, the Fourier transform of a Schwartz function is a Schwartz function, and it is the only set of functions with this property. This is not, in retrospect, a nice property that would be convenient for physicists. The theory of tempered distribution, which allows us to define the Dirac $\delta$-function, is mathematically based, at its core, over the Schwartz space. Therefore, this space already plays a fundamental role in the definitions of the tools we already use.

On the downside, many of the tools we use indeed assume a Hilbert space and many theorems would not hold without Cauchy completeness. There are two responses to this issue. First of all, there is no reason to prohibit that calculation are performed on the larger, unphysical space. When solving an integral, we do sometimes extend the domain for real to complex values, which are unphysical, because complex analysis has nicer mathematical features. The result so found, however, is still the valid result on the real domain. Similarly, we can pose the problem on the Schwartz space, extend to the Hilbert space for calculation, and then bring the result back to the Schwart space. However, there is a second way to look at this issue. Given that the focus has been squarely on Hilbert spaces, we do not know which mathematical results will hold, or even whether other results may be found. This is not something that the authors of this paper can answer and neither the mathematicians that were asked. This requires a serious research effort on the mathematical underpinning of multiple areas, which brings us to our closing point.

This effort is never going to happen if the physics community, as whole, do not generate enough interest on the subject. There may be a better series of mathematical tools and results that stem from a more precise specification of our physical requirements in terms of mathematical definitions. It is unlikely that the mathematical world will give us these tools unprompted. It is the job of us physicists to understand exactly what are ramifications of the definitions in terms of what physical objects would be left out or what unphysical objects would be brought in.

% TODO: hint that the unitary transformations on Schwartz space are a subset of the ones on the Hilbert, and therefore they have unique inverse; operators we are interested in are uniquely extended to the Hilbert space.

>>>>

We have found that:
\begin{enumerate}
	\item complex inner product spaces are needed
	\item infinite dimensional spaces are needed
	\item completeness is unphysical.
\end{enumerate}
In other words, we need a stricter way to characterize convergence within the state space so that we can manage the infinite dimensionality. Conceptually, we need to treat the infinities coming from unbound quantities as potential infinities, not actual infinities: we never actually have infinitely many particles or a system positioned at infinity, we have an infinite range of possible values.

A naive idea would be to require that the expectation value for all possible variables converge. Unfortunately, this doesn't work. Suppose we have a distribution $\rho(x)$ over position $x$. We would be requiring the expectation of any function to be finite, which would include $e^x$, $e^{e^x}$, $e^{e^{e^x}}$ and so on. The only way this can work is if the function has compact support. Physically, this seems too restrictive as it excludes functions like normal distributions. At a closer investigation, the Paley-Wiener theorem tells us that the Fourier transform of a function with compact support is an entire function. The set of zeros of an entire function is either discrete or the full space. Therefore, we cannot have compact support in both position and momentum.

The only way out, then, is to require only a certain class of observables to have finite expectation. Physically, it would make sense to require exactly those observables that are needed to fully reconstruct the state through quantum tomography. A good set in that respect seems to be the set of polynomial of conjugate quantities, like position and momentum. In quantum mechanics, this means requiring all expectations of $X^nP^m = -\imath \hbar x^n\partial_x ^m$ to be finite. This restricts our space to those functions that are infinitely smooth (i.e. all derivative exist) and decrease very rapidly as $x$ goes to infinity: the Schwartz functions. Requiring finite expectation for all polynomials of position and momentum, then, does not give us the usual $L^2$ Hilbert space, it gives us the Schwartz space.

Not only the expectations of all polynomials exist, but they are enough to identify the elements and to define limits. The Schwartz space is in fact a Fr\'echet, meaning that is a Hausdorff space, with the topology induced by the seminorms given by the expectations and it is complete with respect to those seminorms. Physically, this means that measurements on the polynomials (i.e. the topology of the seminorms) are enough to identify the states (i.e. the singletons, which are closed sets as per Hausdorff). Moreover, if we have a sequence for which all expectations converge, we are guaranteed to converge to another state (i.e. completeness with respect to the seminorms). If any expectation diverges, the limit is not in the space. This criteria of convergence seems a lot more physically sound.

To compare Schwartz and Hilbert spaces, the Schwartz space is a subspace of the $L^2$ Hilbert space taken as an inner product space. That is, all Schwartz functions are elements of the Hilbert space, they are closed under linear combinations and have the same inner product. It is not complete and it is a dense subspace, meaning that any element of the Hilbert space can always be approximated by a Schwartz function with arbitrary level of precision. Or, equivalently, that once a map is defined over the Schwartz space, the map is uniquely extended over the Hilbert space. It would seem, then, that all we are losing is the behavior at  actual infinity.

The Schwartz space has also a very important property: it is the only space that is closed under Fourier transform. That is, the Fourier transform of a Schwartz function is a Schwartz function, and it is the only set of functions with this property. Given the importance of the Fourier transform in quantum theory, this feels like an important property to have. Mathematically, the theory of tempered distribution, which allows us to define objects like the Dirac $\delta$-function, is built on top of the Schwartz space. Therefore, this space already plays a fundamental role in the definitions of the tools we already use.

If we restrict ourselves to the Schwartz space, then, the only unitary transformations we are interested in are those that map Schwartz function to Schwartz functions. These are special cases of unitary transformation over Hilbert spaces, and inherit all their properties, such as the existence of the inverse. The change of coordinate we defined in the previous section, then, is automatically ruled out. It is interesting to note that two Hilbert spaces connected by such unitary transformations have essentially have two separate theories of distributions as they start from two different sets of objects. In our mind, this shows exactly how two inequivalent physical representations can be mathematically equivalent. Our suspicion is that the problem of inequivalent mathematical representations in QFT is a consequence of adding another non-physical infinity.

Though we believe the arguments to be thoroughly convincing, we are stopping short from declaring Schwartz spaces physical. First, because this is not the goal of the paper, which is just to show the unphysicality of Hilbert spaces. Second, because we would need to show the importance of expectation values of polynomials from experimental considerations. It is not clear to us that that's the case, so it may be that the justification would have to proceed in a different, but ultimately equivalent, way.

Now, one may be concerned that if we abandon Hilbert spaces we would also have to abandon many convenient tools built on top. This is not the case. When solving an integral, we do sometimes extend the domain for real to complex values, which are unphysical, because complex analysis has nicer mathematical features. The result so found, however, is still the valid result on the real domain. Similarly, we can pose the problem on the Schwartz space, extend to the Hilbert space or to the space of distributions for calculation, and then bring the result back to the Schwartz space. What really need to be rethought, instead, is the construction of state space, especially under infinite operations. If we want to construct the state space for a composite system of potentially infinitely many subsystems, understanding exactly what infinities are physical and which aren't is exactly the issue. Therefore taking the Hilbert space, which we saw doesn't deal with physical infinities correctly, and making infinitely many copies of it is bound to create problems.
